The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-14 paper.tex
Motivation - type providers, AI assistants

\bibliographystyle{plainurl}
\title{The Choose-Your-Own-Adventure Calculus} % (Pearl)

@@

The main contribution of this paper is conceptual rather than technical. Programming language
researchers have long focused on \emph{programming languages} rather than on a more general notion
of interactive \emph{programming systems} \cite{rpg-2012-revolution,jakubovic-2023-techdims}.
However, programming language theory methods can be equally useful for studying interactive
systems~\cite{adams-2025-grove}. The choose-your-own-adventure-calculus is another step in this
direction. It brings to light a pattern that is perhaps not surprising, but is very easy to
overlook until it is given a name.

@@

@@

\cite{jakubovic-2023-techdims}
\cite{petricek-2020-live}

>>>>

\newcommand{\ident}[1]{\textsf{#1}}

\bibliographystyle{plainurl}
\title{The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)}

@@

The main contribution of this paper is conceptual rather than technical. We capture a
pattern that is perhaps not surprising in retrospect, but that is easy to overlook until it is
given a name. We use formal programming language theory methods to precisely describe
interesting aspects of the pattern. Moreover, our work also confirms that programming
language theory methods can be extremely effective for studying not just \emph{programming
languages}, but also interactive \emph{programming systems}~\cite{jakubovic-2023-techdims}.

@@

Computer scientists studying programming have long focused on programming languages as syntactic
entities, sometimes neglecting the interactive environments in which the languages are inevitably
embedded \cite{rpg-2012-revolution}. Notably, in many of the motivating examples that we draw
from in this section, the interactive aspect of the system is only described in supplementary
materials such as tutorial notes \cite{brady-2015-idris}, workshop papers \cite{syme-2013-inforich}
or user guides \cite{altenkirch-1994-alf}. Only recently, programming language theory has been used
to look at interactive environments \cite{adams-2025-grove,mayer-2018-bidirectional}. Our work aims
to contribute to this research direction.

The following four sections briefly review four different instances of the choose-your-own-adventure
interaction pattern. In all of those, an interactive editor offers the user some kind of a
completion list during working with the system.

\begin{figure}[t]
\includegraphics[width=0.7\textwidth]{fig/worldbank.png}
\caption{F\# code editor showing completions offered by the World Bank type provider.}
\label{fig:worldbank}
\end{figure}

\subparagraph{Type providers}

F\# type providers \cite{syme-2013-inforich} are a mechanism that make it possible to integrate
external data sources into the F\# type system. A type provider is a compiler extension, loaded and
executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
external data and use it to generate a suitable statically-typed representation of the external
data, typically as objects with members. Type providers can, for example, infer the type from a
sample JSON \cite{petricek-2016-fsdata} or read a database schema.

The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
from the World Development Indicators database. The provided \ident{wb} object allows the programmer
to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
and an \ident{[Indicator]} in a chain of members
\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
The result is a time series with values of the given indicator and country. More generally,
the example can be seen as a special case of a type provider for slicing n-dimensional
data cube \cite{petricek-2022-thegamma}. Here, we choose a fixed value for two of the three
dimensions (country, indicator, time).

When using the type provider, the user types the first line of code and trigger auto-completion
by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
option from a list and typing another dot.\footnote{This interaction pattern has been
lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}

\begin{figure}[t]
\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
each athlete and sort the data by the number of gold medals.}
\label{fig:thegamma}
\end{figure}

\subparagraph{Data exploration}

The Gamma~\cite{petricek-2022-thegamma} is a web-based programmatic data exploration environment
for non-programmers. In The Gamma, type providers are the primary programming mechanism and they
are used not just for data access, but also for constructing queries.

The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
of the data schema and uses the schema to generate all possible valid parameters. When sorting
data, it generates an object with two members for each columns -- one for ascending and one for
descending sort. Similarly, grouping first offers all columns as possible sorting keys and then
lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
The system evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.

The interaction pattern is the same as above. The user triggers auto-completion by entering a
data source and typing dot. They can then repeatedly select an operation and its parameters to
construct a query. One notable difference is that the structure of the generated types is
potentially infinite (the user can keep adding further operations) and so the types are generated
lazily.

\subparagraph {AI assistants}

The third instance of the choose-your-own-adventure interaction pattern comes from work on
semi-automatic data wrangling tools known as AI assistants \cite{petricek-2023-aias}.
An AI assistant guides the analyst through a data wrangling problem such as reconciling mismatched
datasets, filling missing values or inferring data format and types. An AI assistant tries to solve
the problem automatically and suggests an initial data transformation, but it also generates a
number of constraints that the user can choose from to refine the initial solution. If the initial
solution is not correct, the user chooses a constraint and the AI assistant runs again, suggesting
a new data transformation that respects the constraint.

Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
quality data published by Ofcom for two subsequent years. The format of the CSV files differs.
Some columns were added, some removed, all were renamed and their order has changed. In the
example, we selected 6 columns from the year 2015 and want to find matching data from 2014.

When the AI assistants runs automatically, it incorrectly maps the \ident{Urban.rural} (2014)
column to \ident{Nation} (2015). This happens because both columns are categorical and have three
values with similar distribution, but a data analyst can easily spot the mistake. To correct the
error, they click the ``+'' button to add a constraint and choose \ident{Don't match Urban.rural
and Nation} to specify that the two columns should not be matched. Datadiff then runs again and
finds the correct matching.

\begin{figure}[t]
  \includegraphics[width=0.9\textwidth]{fig/aia.png}
  \caption{aia}
  \label{fig:aia}
\end{figure}

The interaction patter is the same as in the previous two cases. The analyst constructs the
correct data transformation by repeatedly choosing from a list of options, until they obtain
the desired result. However, the way the interaction pattern is used in this case differs.
First, in the case of type providers, we are constructing a program (adding operations to a
method chain). Now, we are collecting constraints and the AI assistant then synthesizes a
data transformation (a script) based on those constraints. Second, in the case of type providers,
the completion list offered all possible members of the object. Now, the list offers a choice of
constraints that are recommended by the AI assistant, which may not be complete.

\subparagraph{Interactive theorem provers}

yo

@@

\cite{rein-2019-exploratory}
